### 2010-12-29 Documentation/atropos/atropos.tex
ENH: multivariate atropos example
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@731 fd2da0a4-d968-48df-9cb5-d8c496287cbf

% \doublespacing

@@

\begin{center}
{\Large \bf An Open-Source Multivariate Framework for $n$-Tissue
  Segmentation with Evaluation on Public Data} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Jue Wu$^1$, Philip A.~Cook$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\subsection{Initializing the Atropos Objective}
Atropos has a number of parameters defined within
Listing \ref{listing:n4} and will function on 2, 3 or 4 dimensional
data.  However, the majority of the time, users will be concerned with
a smaller set of input parameters.  Here, we list the recommended input and
an example definition for each parameter:
\begin{description}
\item[Input images to be segmented:] If more than
  one input image is passed, then a multivariate model will be
  instantiated.  E.g.  {\ttfamily -a Image.nii.gz} for one image and
  {\ttfamily -a Image1.nii.gz -a Image2.nii.gz} for multiple images.
\item[Input image mask:] This binary image
  defines the spatial segmentation domain.  Voxels outside the
  masked region are designated with the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
\item[Convergence criteria:] The algorithm terminates if it reaches  the maximum number of iterations or
  produces a change less than the minimum threshold change in the posterior.
  E.g. {\ttfamily -c [5,1.e-5]}. 
\item[MRF prior:] The key parameter to increase or decrease the
  spatial smoothness of the label map is $\beta$.  A useful range of
  $\beta$ values is $0$ to $0.5$ where we usually use 0.05, 0.1 or 0.2
  in brain segmentation.  E.g.  {\ttfamily -m [0.1, 1x1x1]} would
  define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.
\item[Initialization:] The initialization options include (where the
  first parameter defines $K$, here 3 for each below),
\begin{itemize}
\item  {\ttfamily -i  Kmeans[3]} standard K-means initialization for
  three classes,
\item  {\ttfamily -i  PriorLabelImage[3,label\_image.nii.gz]} and
%  where scalar $w \in [0,1]$ is unused. 
\item  {\ttfamily -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,w]} 
where $w=0$~(use the prior probability images only for
  initialization) or $w>0.0$ (use the prior probability images
  throughout the optimization).  If one chooses $0 < w < 1.0$ then one
  will increase (from zero) the weight on the priors.  These images,
  like the {\ttfamily PriorLabelImage}, should be defined with the same domain as
  the input images to be segmented.  
\end{itemize}
\item[Posterior formulation:]  The user may choose to estimate the
  mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
  selecting the {\ttfamily PriorLabelImage} initialization and   {\ttfamily -p Plato[0]}.
\item[Output:] Atropos will output the hard segmentation and the
  probability image for each model.  E.g.  {\ttfamily -o
    [segmentation.nii.gz,seg\_prob\%02d.nii.gz]} will write
  out the hard segmentation in the first output parameter and a
  probability image for each class named, here, seg\_prob01.nii.gz,
  seg\_prob02.nii.gz, etc. 
\end{description}
Higher dimensions than 4 are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same
dimension, size, origin, etc.), will automatically enable multivariate
likelihoods.  In that case, the first image specified on the 
command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
follow-up multivariate K-means labeling.    More details on each of
the key implementation options are given below.

@@

\subsection{Integration with Registration and N4 Bias Correction} 
Much of the image processing in our lab takes advantage of multiple
open source tools.  Most critically, these involve registration and
bias correction, in addition to segmentation. 
\subsubsection{Registration of Probability Maps} 
Image registration enables the ability to transform information
between spatial domains which may aid in both segmentation and bias
correction.  We rely heavily on template-building strategies
\citep{Avants2010,Avants2010a} which are also offered in ANTs.  Since
aligned prior probability images and prior label maps are often
associated with such templates, Atropos can be initialized with these
data with their influence regulated by a prior probability weighting
term.  Although prior label maps can be specified as a single
multi-label image, prior probability data are often represented as
multiple scalar images with a single image corresponding to a
particular label.  For relatively small classifications, such as the
standard 3-tissue segmentation (i.e. white matter, gray matter, and
cerebrospinal fluid), this does not typically present computational
complexities using modern hardware.  However, when considering dense
cortical parcellations where the number of labels can range upwards of
74 per hemisphere \citep{Destrieux2010}, the memory load can be
prohibitive if all label images are loaded into run-time memory
simultaneously.  A major part of minimizing memory usage in Atropos,
which corresponds to the boolean `{\ttfamily
-u/--minimize-memory-usage}' option, is the sparse representation of
each of the prior probability images.  Motivated by the observation
that these spatial prior probability maps tend to be highly localized
for large quantities of cortical labels, a threshold is specified on
the command line (default = 0.0) and only those probability values
which exceed that threshold are stored in the sparse representation.
During the course of optimization, the prior probability image for a
given label is reconstructed on the fly as needed.  For instance, the
NIREP (www.nirep.org) evaluation images are on the order of $300
\times 300 \times 256$ with 32 cortical labels.  Our novel memory
minimizing image representation typically shrinks run-time memory usage from a
peak of 10+ GB to approximately 1.5 GB and enable these datasets to be
used for training/prior-based cortical parcellation.

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\subsubsection{Integration of N4 Bias Correction}
The typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using the recently developed N4
algorithm \citep{Tustison2010} which is based on the popular
non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight Toolkit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.  An example of this procedure, using BrainWeb data with
40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
information necessary to repeat this example in the script entitled
`{\ttfamily  atroposBwebRF40FigureExample.sh}' which is available in the ANTs toolkit's Atropos
documentation folder as of svn commit 711.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=6in]{Figures/bwebN4ex.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
sequential processing and apply to BrainWeb single-subject data with
40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
$0.2$.  Slice 71 of the input data is in (a).  The initial K-means
($K=3$) segmentation is in panel (b).  We use the brain mask to guide
N4 bias correction and produce the image, shown in (c).  We repeat the
K-means segmentation, but with the N4 corrected image as input, and
produce the segmentation in (d).  The average 3-tissue Dice overlap of
result (b) is 0.906 while the average overlap for (d) is 0.954.
Arrows highlight a region of large before-after segmentation
discrepancy.}
\label{fig:bwebrf40}
\end{center}
\end{figure}

@@

bias level.

@@

\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}

@@

\caption{\baselineskip 12pt \small BrainWeb single-subject results for
  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  The prior label image
  initialization is competitive with but inferior to K-means.  The
  legend is in the same position in each graph, allowing a visual
  comparison of the results.  As one may see, the N4 assisted overlap
  values are consistent across bias field/RF inhomogeneity.}
\label{fig:bweb1}

@@

\subsubsection{20 Subject Evaluation} The single-subject BrainWeb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject BrainWeb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=1$, the priors are only maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}

\subsection{The Hammers Dataset Evaluation}
\label{sec:hammers}
We evaluate the ability to improve multi-template labeling results by
converting the group labels to probability maps and using them to
drive many-class EM segmentation.  The ground truth labels cover 69
classes and much of the brain.  Some unlabeled regions remain which we
assign to label 69 such that all brain parenchyma contains a unique
label.  Following \cite{Avants2010a}, the initialization of our
evaluation applies the script {\verb ants_multitemplate_labeling.sh }
(available in the ANTs toolkit) to the 19 Hammers evaluation datasets
located at \url{http://www.brain-development.org/}
\citep{Hammers2003,Heckemann2006}.  These initial majority voting
results are competitive with prior work
\citep{Heckemann2006,Heckemann2010} and serve as a baseline against
which we compare.

@@

\subsection{Reproducibility of this Evaluation}
The BrainWeb data is freely available.  We used single-subject
BrainWeb data as is but added a meta format data header to the raw
binary files.   An example copy of this header is contained in {\verb atroposBwebRF40FigureExample.sh }.  The 20 subject data, however, required excluding
non-cerebrum tissue classes.  The Hammers data was also used as is.
The scripts for the single-subject BrainWeb study is available in
supplementary material.  The initialization procedure for the BrainWeb 20 and
the Hammers evaluation data is based on freely available scripts
included with the ANTs toolkit, specifically  {\verb ants_multitemplate_labeling.sh } and  {\verb atroposBwebRF40FigureExample.sh }.

@@

\subsection{Clinically-Related Evaluation}
While specifying performance on BrainWeb is highly
valuable, clinical validation is a second important aspect of
segmentation evaluation.  For instance, 
\citep{Freeborough1997,Westlye2009,Sanchez-Benavides2010,Chou2009,Bresser2011}
are only a few of the papers that evaluate segmentation performance with
respect to a known neurobiological outcome measure.  Atropos is
currently used in clinical studies and a number of clinically-focused,
application-specific evaluations are ongoing and will constitute
future work.  One early example of a clinically-focused Atropos
neuroimaging application is in \citep{Avants2010c}.  An additional successful
application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
\citep{Tustison2010a} which also used the open source Glamorous Glue
algorithm to impose topology constraints \cite{Tustison2010b}.  Thus, future work may
incorporate topology more closely into the Atropos methodology.

@@

>>>>

\doublespacing

@@

\begin{center}
{\Large \bf An Open Source Multivariate Framework for $n$-Tissue
  Segmentation with Evaluation on Public Data} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Jue Wu$^1$, Philip A.~Cook$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\doublespacing
\subsection{Initializing the Atropos Objective}
Atropos has a number of parameters defined within
Listing \ref{listing:n4} and will function on 2, 3 or 4 dimensional
data.  However, the majority of the time, users will be concerned with
a smaller set of input parameters.  Here, we list the recommended input and
an example definition for each parameter:
\begin{description}
\item[Input images to be segmented:] If more than
  one input image is passed, then a multivariate model will be
  instantiated.  E.g.  {\ttfamily -a Image.nii.gz} for one image and
  {\ttfamily -a Image1.nii.gz -a Image2.nii.gz} for multiple images.
\item[Input image mask:] This binary image
  defines the spatial segmentation domain.  Voxels outside the
  masked region are designated with the label $0$.  E.g.  {\ttfamily -x mask.nii.gz}. 
\item[Convergence criteria:] The algorithm terminates if it reaches  the maximum number of iterations or
  produces a change less than the minimum threshold change in the posterior.
  E.g. {\ttfamily -c [5,1.e-5]}. 
\item[MRF prior:] The key parameter to increase or decrease the
  spatial smoothness of the label map is $\beta$.  A useful range of
  $\beta$ values is $0$ to $0.5$ where we usually use 0.05, 0.1 or 0.2
  in brain segmentation.  E.g.  {\ttfamily -m [0.1, 1x1x1]} would
  define $\beta=0.1$ with a MRF radius of one voxel in each of three dimensions.
\item[Initialization:] The initialization options include (where the
  first parameter defines $K$, here 3 for each below),
\begin{itemize}
\item  {\ttfamily -i  Kmeans[3]} standard K-means initialization for
  three classes,
\item  {\ttfamily -i  PriorLabelImage[3,label\_image.nii.gz]} and
%  where scalar $w \in [0,1]$ is unused. 
\item  {\ttfamily -i  PriorProbabilityImages[3,label\_prob\%02d.nii.gz,w]} 
where $w=0$~(use the prior probability images only for
  initialization) or $w>0.0$ (use the prior probability images
  throughout the optimization).  If one chooses $0 < w < 1.0$ then one
  will increase (from zero) the weight on the priors.  These images,
  like the {\ttfamily PriorLabelImage}, should be defined with the same domain as
  the input images to be segmented.  
\end{itemize}
\item[Posterior formulation:]  The user may choose to estimate the
  mixture proportions (or not) by setting  {\ttfamily -p Socrates[1]}
  or {\ttfamily -p Socrates[0] }.  Fixed label boundary conditions may be employed by
  selecting the {\ttfamily PriorLabelImage} initialization and   {\ttfamily -p Plato[0]}.
\item[Output:] Atropos will output the hard segmentation and the
  probability image for each model.  E.g.  {\ttfamily -o
    [segmentation.nii.gz,seg\_prob\%02d.nii.gz]} will write
  out the hard segmentation in the first output parameter and a
  probability image for each class named, here, seg\_prob01.nii.gz,
  seg\_prob02.nii.gz, etc. 
\end{description}
Higher dimensions than 4 are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same
dimension, size, origin, etc.), will automatically enable multivariate
likelihoods.  In that case, the first image specified on the 
command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
follow-up multivariate K-means labeling.    More details on each of
the key implementation options are given below.

@@

\doublespacing
\subsection{Integration with Registration and N4 Bias Correction} 
Much of the image processing in our lab takes advantage of multiple
open source tools.  Most critically, these involve registration and
bias correction, in addition to segmentation. 
\subsubsection{Registration of Probability Maps} 
Image registration enables the ability to transform information
between spatial domains which may aid in both segmentation and bias
correction.  We rely heavily on template-building strategies
\citep{Avants2010,Avants2010a} which are also offered in ANTs.  Since
aligned prior probability images and prior label maps are often
associated with such templates, Atropos can be initialized with these
data with their influence regulated by a prior probability weighting
term.  Although prior label maps can be specified as a single
multi-label image, prior probability data are often represented as
multiple scalar images with a single image corresponding to a
particular label.  For relatively small classifications, such as the
standard 3-tissue segmentation (i.e. white matter, gray matter, and
cerebrospinal fluid), this does not typically present computational
complexities using modern hardware.  However, when considering dense
cortical parcellations where the number of labels can range upwards of
74 per hemisphere \citep{Destrieux2010}, the memory load can be
prohibitive if all label images are loaded into run-time memory
simultaneously.  A major part of minimizing memory usage in Atropos,
which corresponds to the boolean `{\ttfamily
-u/--minimize-memory-usage}' option, is the sparse representation of
each of the prior probability images.  Motivated by the observation
that these spatial prior probability maps tend to be highly localized
for large quantities of cortical labels, a threshold is specified on
the command line (default = 0.0) and only those probability values
which exceed that threshold are stored in the sparse representation.
During the course of optimization, the prior probability image for a
given label is reconstructed on the fly as needed.  For instance, the
NIREP (www.nirep.org) evaluation images are on the order of $300
\times 300 \times 256$ with 32 cortical labels.  Our novel memory
minimizing image representation typically shrinks run-time memory usage from a
peak of 10+ GB to approximately 1.5 GB and enable these datasets to be
used for training/prior-based cortical parcellation.

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\doublespacing
\subsubsection{Integrating N4 Bias Correction}
The typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using the recently developed N4
algorithm \citep{Tustison2010} which is based on the popular
non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight Toolkit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.

@@

bias level.  An example of this procedure, using BrainWeb data with 40\% RF bias,
is in figure~\ref{fig:bwebrf40}.  We supply the information necessary
to repeat the results in this figure in the script entitled
`{\ttfamily atroposBwebRF40FigureExample.sh}' which is available in
the ANTs toolkit's Atropos documentation folder as of svn commit 711.
The script may be easily modified to run the whole evaluation.  In
addition, Figure~\ref{fig:bwebrf40} shows the results of simultaneously using proton density
and T1 weighted BrainWeb data to perform the segmentation.  This
multivariate input data outperforms the univariate T1 weighted data
alone. 

@@

\includegraphics[width=6in]{Figures/bwebN4ex.pdf}

@@

\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
sequential processing and apply to BrainWeb T1-weighted single-subject
data with 40\% RF bias and 3\% noise.  The $\beta$ for the MRF term
is, here, $0.2$.  Slice 71 of the input data is in (a).  The initial
K-means ($K=3$) segmentation is in panel (b).  We use the brain mask
to guide N4 bias correction and produce the image in (c).  We
repeat the K-means segmentation, but with the N4 corrected image as
input and produce the segmentation in (d).  The average 3-tissue Dice
overlap of result (b) is 0.906 while the average overlap for (d) is
0.954.  Arrows highlight a region of large before-after segmentation
discrepancy.  In (e) we see the BrainWeb proton density image with no
inhomogeneity and 3\% noise.  Its segmentation is in (f) with average
3-tissue Dice overlap of 0.895.  In (g) we use both proton density
data and T1 data as multiple modality input to Atropos.  The segmentation
of this two-modality input data, using a multivariate Gaussian model,
produces average 3-tissue Dice overlap of 0.958, which exceeds
the univariate solution.  An arrow highlights one region where there
is small, visually recognizable improvement in sulcal segmentation
relative to the result from T1 data alone.  A second area of
improvement is the putamen segmentation.  The ground truth
segmentation is in (h).  The multivariate segmentation result, in
combination with the low PD segmentation performance, suggests PD and
T1 provide complementary information that may improve 3-tissue
segmentation and serves to validate the multivariate Atropos
implementation.
\label{fig:bwebrf40}

@@

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_B_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_N4_RF40.pdf}\\
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF0.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF20.pdf}
\includegraphics[height=2in]{Figures/tissue_dice_vs_mrf_PLI_RF40.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb single-subject results for
  each tissue.  The results show that N4 bias correction, combined with Atropos, results in a minimal effect of
  bias, even at the 40\% level.  The optimal $\beta$ for the MRF term
  appears to be between $0.1$ and $0.2$.  The prior label image
  initialization is competitive with but inferior to K-means.  The
  legend is in the same position in each graph, allowing a visual
  comparison of the results.  As one may see, the N4 assisted overlap
  values are consistent across bias field/RF inhomogeneity.}
\label{fig:bweb1}
\end{center}
\end{figure}
\subsubsection{20 Subject Evaluation} The single-subject BrainWeb
study in the previous section tested the basic Atropos options and the
benefit of N4 for segmentation in the presence of bias.  The 20
subject BrainWeb data allows us to use 2-fold cross-validation to test
our ability to segment different individuals reliably.  In this study,
we divide the 20 subjects equally into training and testing groups.
We then exploit the ground-truth labeling of the training data to
build both a group template \citep{Avants2010} and also prior
probability maps for each of the three major tissues in the cerebrum.
Each prior probability map is gained by deforming the ground truth
labels from each of the 10 training subjects to their template and
averaging component by component.  We then deform the template---and
priors---to the ten testing subjects and run Atropos with not only
{\ttfamily KMeans[3]} initialization but also {\ttfamily
  PriorProbabilityMap[3,priors\%02d.nii.gz,$w$] } where $w \in
\{0.0,0.5\}$.  We then switch the roles of testing and training sets to
gain 3-tissue segmentation for each of the twenty subjects.  
When $w=0$, the priors are only used in initializing
the model parameters but not during subsequent EM iterations.  
When $w=1$, the priors are only maintained in the product with the
likelihood during all EM iterations.  Results, in terms of bar
plots for Dice overlap mean and standard deviation, are shown in figure~\ref{fig:bweb20}.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=2.2in]{Figures/bw20_kmeans.png}
\includegraphics[width=2.2in]{Figures/bw20_priorIni.png}
\includegraphics[width=2.2in]{Figures/bw20_priorSeg.png}
\end{tabular}
\caption{\baselineskip 12pt \small BrainWeb 20-subject results for
  each tissue as a function of MRF-$\beta$ parameter where MRF-$\beta$
  is in $\{0,0.05,0.1,0.15,0.2\}$ and increases left to right.  The results show that the PriorProbabilityMaps with
  $w=0.5$ (far right) gives the best performance for all tissues.}
\label{fig:bweb20}
\end{center}
\end{figure}

\subsection{The Hammers Dataset Evaluation}
\label{sec:hammers}
We evaluate the ability to improve multi-template labeling results by
converting the group labels to probability maps and using them to
drive many-class EM segmentation.  The ground truth labels cover 69
classes and much of the brain.  Some unlabeled regions remain which we
assign to label 69 such that all brain parenchyma contains a unique
label.  Following \cite{Avants2010a}, the initialization of our
evaluation applies the script {\ttfamily ants\_multitemplate\_labeling.sh }
(available in the ANTs toolkit) to the 19 Hammers evaluation datasets
located at \url{http://www.brain-development.org/}
\citep{Hammers2003,Heckemann2006}.  These initial majority voting
results are competitive with prior work
\citep{Heckemann2006,Heckemann2010} and serve as a baseline against
which we compare.

@@

\subsection{Reproducibility of this Evaluation}
\noindent{\bf Data:} The BrainWeb data is freely available.  We used
single-subject BrainWeb data as is but added a meta format data header
to the raw binary files.  An example copy of this header is contained
in {\ttfamily atroposBwebRF40FigureExample.sh }.  The 20 subject data,
however, required excluding non-cerebrum tissue classes.  The Hammers
data was also used as is (\url{http://www.brain-development.org/}).

\noindent{\bf Software:} The ANTs software is available at \url{http://www.picsl.upenn.edu/ANTs} with download and compilation
instructions at \url{http://www.picsl.upenn.edu/ANTS/download.php}.
Svn release 711 was used for the examples and evaluations performed in
this paper.  Some components of ANTs depend on the Insight ToolKit.
The most critical dependency, for Atropos, is the ITK statistics framework
used to implement the univariate and multivariate parametric models.
We linked to the git version of ITK current as of Dec. 1, 2010.  See
\url{http://www.itk.org/Wiki/ITK/Git/Download} for instructions on git ITK.

\noindent{\bf Scripts:} The complete script for the single-subject
BrainWeb study is based on generalizing {\ttfamily atroposBwebRF40FigureExample.sh }, which is available in the ANTs
toolkit (svn release 711 or greater) and which reproduces Figure 2
results.  The template-based normalization procedure for the BrainWeb
20 and the Hammers evaluation data is based on freely available
scripts included with the ANTs toolkit, specifically {\ttfamily
  ants\_multitemplate\_labeling.sh } and {\ttfamily
  buildtemplateparallel.sh }.  A release version of ANTs---with a final version of Atropos---will
be prepared with the final version of this manuscript.

@@

\subsection{Clinically-Related Evaluation}
While specifying performance on BrainWeb is highly
valuable, clinical validation is a second important aspect of
segmentation evaluation.  For instance, 
\citep{Freeborough1997,Westlye2009,Sanchez-Benavides2010,Chou2009,Bresser2011}
are only a few of the papers that evaluate segmentation performance with
respect to a known neurobiological outcome measure.  Atropos is
currently used in clinical studies and a number of clinically-focused,
application-specific evaluations are ongoing and will constitute
future work.  One early example of a clinically-focused Atropos
neuroimaging application is in \citep{Avants2010c}.  An additional successful
application area is that of ventilation-based segmentation of hyperpolarized helium-3 MRI
\citep{Tustison2010a} which also used the open source Glamorous Glue
algorithm to impose topology constraints \citep{Tustison2010b}.  Thus, future work may
incorporate topology more closely into the Atropos methodology.

@@

%# T1_RF40_3%_N4 reference results 
 Label 1 DICE 0.941891  RO 0.890165 TP1 0.904786 TP2 0.98217
 Label 3 DICE 0.966776  RO 0.935688 TP1 0.962814 TP2 0.97077
 Label 2 DICE 0.953862  RO 0.911794 TP1 0.971043 TP2 0.937279
 AvgDice  : 0.954176
%# T1_RF40_3%_N4 + T2_RF0_3% noise reference results 
Label 1 DICE 0.937508  RO 0.882367 TP1 0.889406 TP2 0.99111
 Label 3 DICE 0.96701  RO 0.936128 TP1 0.958398 TP2 0.975778
 Label 2 DICE 0.951639  RO 0.90774 TP1 0.977645 TP2 0.926981
 AvgDice  : 0.952052
%# T1_RF40_3%_N4 + PD_RF0_3% noise reference results 
 Label 1 DICE 0.944732  RO 0.895254 TP1 0.912413 TP2 0.979425
 Label 3 DICE 0.970348  RO 0.942405 TP1 0.969169 TP2 0.971531
 Label 2 DICE 0.957976  RO 0.919341 TP1 0.97094 TP2 0.945353
 AvgDice  : 0.957685
%# PD_RF0_3% noise reference results 
 Label 2 DICE 0.905959  RO 0.828085 TP1 0.892396 TP2 0.919941
 Label 3 DICE 0.942477  RO 0.891212 TP1 0.943874 TP2 0.941084
 Label 1 DICE 0.837847  RO 0.720944 TP1 0.87262 TP2 0.805739
 AvgDice  : 0.895428
%# T2 
 Label 1 DICE 0.931616  RO 0.871986 TP1 0.984176 TP2 0.884385
 Label 2 DICE 0.882997  RO 0.790505 TP1 0.79822 TP2 0.987921
 Label 3 DICE 0.901592  RO 0.820817 TP1 0.99405 TP2 0.82487
 AvgDice  : 0.905402
